Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite

### 2022-04-05 manuscript_v11.Rmd
final ms edits

-   The BIKER (Bayesian Inference of the $k_{600}$ Exchange Rate) algorithm predicts gas exchange velocity and $CO_2$ fluxes solely from simulated SWOT data without calibration
-   BIKER is robust to SWOT measurement errors
-   BIKER will allow for novel study of gas exchange spatiotemporal dynamics after SWOT's launch

@@

Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite can provide new insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established inversion techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Similar to remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%). Finally, BIKER is robust to measurement errors implicit in the SWOT data. Upon SWOT’s launch, we suggest that BIKER will be useful in mapping global-scale fluvial gas exchange and improving CO2 emissions estimates when coupled with river CO2 models.

@@

Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and near-surface turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [@liuImportanceHydrologyRouting2022; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.67%- @liuImportanceHydrologyRouting2022], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

River $CO_2$ evasion is increasingly better constrained and is clearly a critical component of the global carbon cycle. Equation 1 represents this riverine flux given $\delta CO_2$ (the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used throughout this study.

$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$

@@

The structure of equation 1 necessitates that calculations of $FCO_2$ are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@hallGasExchangeStreams2020; @wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number (*Sc*) of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers each increase gas exchange rates [@hallGasExchangeStreams2020]. Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models focus solely on physical explanations for variation in *k* [@hallGasExchangeStreams2020].

Applying these *k* models across watersheds, regions, or continents is called 'upscaling.' This upscaling allows for estimating the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

@@

Global upscalng has been performed in the literature. @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. More recently, @liuImportanceHydrologyRouting2022 performed a first upscaling assessment of monthly temporal dynamics in global river $FCO_2$, though they relied on monthly modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this. In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales.

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23--349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $k_{O_2}$ must have been to produce their 'observations'. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity. Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

We have established that literature has a reasonably good understanding of $\delta CO_2$ and a relatively poorer understanding of $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time. In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly influence fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

@@

This paper is split into two distinct parts: gas-exchange theory/model development (section 2) and BIKER algorithm development/validation (section 3). Section 3 is fundamentally dependent on the results presented in section 2, therefore section 2 presents both theory and results. Both sections detail the data used. We conclude with a discussion (section 4) that encompasses both gas exchange theory and remote sensing. Figure 1 conceptually maps out the algorithm's approach to inferring $k_{600}$ from SWOT data.

![Figure 1: Conceptual overview of the BIKER algorithm. 1: SWOT will directly measure water surface width and elevation (and thus slope) at a series of cross-sections within mass-conserved river reaches. 2: These hydraulics measurements are used to calculate turbulent dissipation in the river channel (section 3). 3: Turbulent dissipation is used to infer $k_{600}$ via a process-based model (section 2).](cache/maps/diagram.tif)

@@

To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for $k_{600}$, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test. Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (section 2.3). We then empirically validate the model (section 2.3)

@@

In addition to hydraulics measured alongside $k_{600}$ and reported above, we expand our dataset of stream hydraulics using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS. This dataset is used to calculate how frequently SWOT observable rivers meet our large-river hydraulic assumptions (section 2.2).

@@

*k* scales with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020], and extensive field and laboratory experiments have converged on the 'small-eddy model' for *k* as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006a; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. The small-eddy model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$ Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean [e.g. @brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface. Further, to our knowledge, this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$ Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall (equation 5). Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

@@

Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953]. This is a common assumption in hydraulic and geomorphic modeling of large rivers. For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of only `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

We therefore assume that all SWOT-observable rivers are hydraulically-wide to derive a model for gas exchange. The overall goal is to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT. To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$). Equation 7 thus defines gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

@@

We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6. While the complete model derivations and results for all four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented and used below.

@@

With equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$. Equation 7 is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

@@

Figure 2 shows that equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well. While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured, remembering that Figure 2 is plotted in logarithmic space. Compared to other combinations of equations 3-6 (Figure S1), there is also less bias in the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows (as expected). The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion built with SWOT in mind. The river hydraulics terms in equation 7 ($\bar{U}$, $H$, and $S$) can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

## 3 Exploiting equation 7 to remotely sense gas exchange velocity

>>>>

-   The BIKER algorithm predicts gas exchange velocity and $CO_2$ fluxes solely from simulated SWOT data without calibration
-   BIKER is robust to SWOT measurement errors
-   BIKER will allow for novel study of gas exchange spatiotemporal dynamics after SWOT's launch

@@

Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/USKA/CSA SWOT satellite can provide new insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation across 166 field measurement of $k_{600}$. We then couple this model with established inversion techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Similar to remote sensing algorithms for SWOT, BIKER is better at explaining the temporal variation of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than reproducing the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%). Finally, BIKER is robust to measurement errors implicit in the SWOT data. Upon SWOT’s launch, we suggest that BIKER will be useful in mapping global-scale fluvial gas exchange and improving $CO_2$ emissions estimates when coupled with river $CO_2$ models.

@@

Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by diffusion and near-surface turbulence at th air/water interface), in addition to their long understood role in transporting carbon to the oceans via downstream advection. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [@liuImportanceHydrologyRouting2022; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' small percentage of the global land surface [0.67%- @liuImportanceHydrologyRouting2022], this carbon flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

River $CO_2$ evasion is increasingly better constrained and is clearly a critical component of the global carbon cycle. Equation 1 represents this riverine flux given $\delta CO_2$ (the air-water flux between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used throughout this study.

$$FCO_2=k \delta CO_2 \mathbf{(1)}$$

@@

The structure of Equation 1 necessitates that calculations of $FCO_2$ are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas flux, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@hallGasExchangeStreams2020; @wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number (*Sc*) of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers each increase gas exchange rates [@hallGasExchangeStreams2020]. Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models focus solely on physical explanations for variation in *k* [@hallGasExchangeStreams2020].

Applying these *k* models across watersheds, regions, or continents is called 'upscaling.' This upscaling allows for estimating the difficult-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

@@

Global upscaling has been performed in the literature. @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. More recently, @liuImportanceHydrologyRouting2022 performed a first upscaling assessment of monthly temporal dynamics in global river $FCO_2$, though they relied on monthly modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this. In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales.

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23--349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $k_{O_2}$ must have been to produce their 'observations'. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity. Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on high fidelity in situ DO data for every river [@hallGasExchangeStreams2020].

We have established that literature has a reasonably good understanding of $\delta CO_2$ and a relatively poorer understanding of $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time. In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via Equation 2) were available at the global scale at a sufficient temporal resolution. Spatially and temporally dynamic hydraulic measurements in turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier.

Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in late 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly influence fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

@@

This paper is split into two distinct parts: gas-exchange theory/model development (Section 2) and BIKER algorithm development/validation (Section 3). Section 3 is fundamentally dependent on the results presented in Section 2, therefore Section 2 presents both theory and results. Both sections detail the data used. We conclude with a discussion (Section 4) that encompasses both gas exchange theory and remote sensing. Figure 1 conceptually maps out the algorithm's approach to inferring $k_{600}$ from SWOT data.

![Figure 1: Conceptual overview of the BIKER algorithm. 1: SWOT will directly measure water surface width and elevation (and thus slope) at a series of cross-sections within mass-conserved river reaches. 2: These hydraulics measurements are used to calculate turbulent dissipation in the river channel (Section 3). 3: Turbulent dissipation is used to infer $k_{600}$ via a process-based model (Section 2).](cache/maps/diagram.tif)

@@

To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for $k_{600}$, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test. Following a description of the data (Section 2.1), we outline established models (Section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (Section 2.3). We then empirically validate the model (Section 2.4)

@@

In addition to hydraulics measured alongside $k_{600}$ and reported above, we expand our dataset of stream hydraulics using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) stream gauge rating curves and made public by the USGS. This dataset is used to calculate how frequently SWOT observable rivers meet our large-river hydraulic assumptions (Section 2.3).

@@

*k* scales with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020], and extensive field and laboratory experiments have converged on the 'small-eddy model' for *k* as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006a; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. The small-eddy model is provided as Equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$ Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation across the entire reach's air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number $Re_*$ formulation. This is Equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean [e.g. @brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface. Further, to our knowledge, this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(Re_*)^{3/8} \mathbf{(4)}$$

Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall (Equation 5). Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (Equation 6) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that Equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

@@

Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT rivers. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953]. This is a common assumption in hydraulic and geomorphic modeling of large rivers. For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of only `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

We therefore assume that all SWOT-observable rivers are hydraulically-wide to derive a model for gas exchange. The overall goal is to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT. To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model (Equation 4) and arrive at Equation 7 (with statistical coefficient $\beta_1$). Equation 7 thus defines gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

@@

We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of Equations 3-4 and Equations 5-6. While the complete model derivations and results for all four models are provided in Text S2 and Figure S1, the final and best-performing model (Equation 7) is presented and used below.

@@

With Equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` measurements of hydraulically-wide $k_{600}$. Equation 7 is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

@@

Figure 2 shows that Equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well. The value for $\beta_1$ is 62.8, which is similar the fitted parameters in other $Re_*$ and $U_*$ models for river *k* [@wangPhysicallyBasedScaling2021]. While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured, remembering that Figure 2 is plotted in logarithmic space. Compared to other combinations of Equations 3-6 (Figure S1), there is also a better fit with the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows (as expected). The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion built with SWOT in mind. The river hydraulics terms in Equation 7 ($\bar{U}$, $H$, and $S$) can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.

## 3 Exploiting Equation 7 to remotely sense gas exchange velocity

We have shown that scaling $k_{600}$ via Equation 7 is useful in hydraulically-wide rivers. Further, Equation 7 has only three non-directly-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @andreadisConstrainingAssimilationSWOT2020; @gleasonGlobalMappingRiver2014; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation. Following a description of the data used (Section 3.1), we detail algorithm development and experimental design (Section 3.2) and then we present the validation results (Section 3.3).
